Death by Dropdown? Engineer Insightful Shiny Apps with Behavioral Science

### 2025-09-09 conference-materials/2025_posit-conf/draft.qmd
Revamp slides with structured narrative and embedded bidux demo
- Rewrote major sections of the conference slides for a more cohesive story
- Consolidated scattered jokes and slide fragments into unified narrative beats
- Embedded actual {bidux} telemetry output and summaries to enhance credibility
- Reframed “dashboard problems” with real-world telemetry evidence and cognitive science tie-ins
- Updated “before vs after” sections to reflect measurable user experience improvements
- Swapped illustrative theory-heavy content for practical, demo-driven walkthroughs

---
title: "Death by Dropdown? Engineer Insightful Shiny Apps with Behavioral Science"
subtitle: "posit::conf(2025)"
author: "Jeremy Winget, PhD"
date: "2025-09-17"
execute:
  echo: true
format:
  revealjs:
    slide-number: true
    transition: fade
    footer: "posit::conf(2025) | jrwinget\\.com"
    preview-links: auto
    code-block-border-left: true
    highlight-style: github
    auto-animate: true
    chalkboard: false
    width: 1600
    height: 900
---

@@

::: {.r-fit-text}
**Ever open a dashboard with 18 filters...**
:::

:::: {.fragment}
::: {.r-fit-text style="color: #e74c3c;"}
**...and immediately close your laptop?**
:::

<div class="tenor-gif-embed" data-postid="17038619" data-share-method="host" data-aspect-ratio="1.77778" data-width="60%"><a href="https://tenor.com/view/ron-swanson-parks-and-recreation-parks-and-rec-computer-dumpster-gif-17038619">Ron Swanson Parks And Recreation GIF</a></div> <script type="text/javascript" async src="https://tenor.com/embed.js"></script>
::::

::: {.notes}
- 15 seconds MAX
- "We've ALL been there"
- Transition: "Quick show of hands"
:::

## Dashboard Support Group 🙋 {.center background-color="#34495e" auto-animate=true}

::: {.incremental}
- Who's built a Shiny app with 10+ filters?
- Who's had a user ask for "just one more dropdown"?
- Who's dashboard has more `selectInput()` than actual insights?
:::

::: {.fragment}
### Welcome to my impromptu support group. You're among friends! 😅

@@

👋 **I'm Jeremy! Behavioral psychologist turned engineer**<br>
*Think of this space as our little therapy session for dropdown overload*
:::

::: {.notes}
- 30 seconds
- Get laughs with "dashboard anonymous"
- "Hi Jeremy" callback
- Transition: "Let me properly introduce myself"
:::

@@

:::: {.columns}
::: {.column width="60%"}
### From Lab to Production

![](img/hex-bidux.png){width="250" .absolute top=150 left=650}

- **PhD in Applied Psychology** <br>*Studied how individuals and groups process information and make decisions*

- **Led Shiny teams at scale** <br>*Built enterprise dashboards that folks actually use*

- **Created {bidux} package** <br>*Your behavioral scientist in the console*

::: {.fragment}
### What is Behavioral Science?
*The study of how people **actually** make decisions (spoiler: not rationally)*
:::
:::

::: {.column width="40%" .fragment}
### My Core Insight

> "Most dashboard failures aren't technical, they're psychological. We're building for computers, not humans."

**Today:** Transform your dashboards using the science of human behavior
:::
::::

::: {.notes}
- 45 seconds
- "Behavioral science = how humans actually work"
- "Not how we think they work"
- Transition: "Let me show you the framework"
:::

## Your Dashboard = A Road Trip 🚗 {background-color="#34495e"}

:::: {.columns .fragment}
::: {.column width="50%"}
### You wouldn't make a long drive without:
- 📍 **I**nterpreting travel guides
- ⚠️ **N**oticing any mechanical problems
- 🌦️ **A**nticipating roadblocks ahead
- 🗺️ **S**tructuring the best route
- ✅ **V**alidating everyone arrived happy
:::

::: {.column width="50%"}
### So why build dashboards without:
- 📍 **I**nterpreting user needs
- ⚠️ **N**oticing friction points
- 🌦️ **A**nticipating behavior
- 🗺️ **S**tructuring a journey
- ✅ **V**alidating understanding
:::

::: {.callout-tip}
### Helpful mnemonic device:
**"Interfaces Need A Smart Vision"**
:::
::::

::: {.notes}
- 20 seconds
- "This is the core idea of the BID framework"
- Transition: "Let me show you the GPS"
:::

## The BID Framework: Your GPS for Dashboard Design

![](img/bid-framework.png)

**Behavioral Insight Design (BID)**: *Not just UX, it's behavioral science in your workflow*

::: {.notes}
- 20 seconds MAX
- "BID framework = your GPS"
- "Today we focus on the middle three - where psychology transforms dashboards"
- "Let's take a look at a example from a client I recently worked with"
:::

## The Problem: 18 Filters, 0 Insights {background-image="img/ui-before.png" background-size="contain" background-position="right"}

::: {.fragment .absolute top=500 left=-40}
### We gathered telemetry on behavior:
- 😵 **65% abandon** in 30 seconds
- ⏱️ **4.3 minutes** to first click
- 🎯 **31% misinterpret** the data
- 📊 **82% use only 3** of 18 filters
:::

::: {.fragment .absolute top=700 left=-40}
**This is behavioral science *screaming* at us**
:::

::: {.notes}
- 45 seconds
- "Let me show you what users actually do"
- Transition to telemetry
:::

## INTERPRET: Plan Your Destination 🗺️ {background-color="#3498db" auto-animate=true}

```{r interpret-stg}
library(dplyr)
library(bidux)

# Define the destination and goals
interpret_stage <- bid_interpret(
  central_question = "Which markets are driving performance?",
  data_story = list(
    hook = "Q4 revenue hit record high, but satisfaction dipped",
    context = "After aggressive marketing across all regions",
    tension = "West region satisfaction fell 10 points",
    resolution = "Focus retention efforts on underperforming regions"
  ),
  user_personas = list(
    list(
      name = "Product Manager",
      goals = "Monitor weekly KPIs",
      pain_points = "Too many filters to find important insights",
      technical_level = "Moderate"
    )
  )
)

summary(interpret_stage)
```

::: {.fragment .callout-tip}
## The 10-Word Guideline:
**Can't explain your dashboard's purpose in 10 words?**

Users won't find answers in 100 filters
:::

::: {.notes}
- 30-45 seconds ONLY
- "A lot of devs think they do this, but telemetry says otherwise"
- "Most jump straight to filters"
- Transition: "But here's what they're missing..."
:::

## NOTICE: The Telemetry Bombshell 💣 {background-color="#e74c3c" auto-animate=true}

:::: {.columns}
::: {.column width="50%"}
![](img/telemetry-heatmap.png)
:::

::: {.column width="50%"}
```{r}
#| eval: false
#| code-line-numbers: "4-5,9"
# Actual production data
bid_telemetry("dashboard.sqlite") |>
  summary()
#> 82% use only 3 filters
#> 4.3 min to first insight
#> 65% abandonment rate

# Let bidux diagnose the problem
bid_notice(telemetry = TRUE)
#> Theory: Cognitive Load Theory
#> Severity: CRITICAL
```
:::
::::

::: {.fragment}
![](img/anakin.jpg){.absolute top=400 right=50 width="400"}
:::

::: {.notes}
- 2 minutes on NOTICE
- "This is your behavioral microscope"
- "Every hover tells a story"
:::

## NOTICE: What 1000 Sessions Reveal {background-color="#e74c3c"}

:::: {.columns}
::: {.column width="60%"}
![](img/telemetry-sessions.png)
:::

::: {.column width="40%"}
### The Hidden Costs
::: {.incremental}
- **Hover time:** 3.2 sec before clicking
- **Help icons:** 68% hover, 0% click
- **Filter #7:** Abandonment spike
- **Errors:** "No results" 319 times
:::

::: {.fragment}
**Your users are drowning in choices**
:::
:::
::::

::: {.notes}
- "Cognitive load is measurable"
- Transition: "Once you spot problems, anticipate how users will cope"
- Transition: "How do we do that?"
:::

## ANTICIPATE: Users Create Their Own Stories 🎭 {background-color="#f39c12"}

::: {.large}
**Users don't see data. They see stories.**
:::

<div class="tenor-gif-embed" data-postid="22483518" data-share-method="host" data-aspect-ratio="1.58416" data-width="50%"><a href="https://tenor.com/view/its-always-sunny-in-philadelphia-charlie-conspiracy-gif-22483518">Charlie Conspiracy GIF</a></div>

::: {.fragment}
**Without guidance, they'll find patterns that don't exist**
:::

::: {.notes}
- Start of 2 minutes on ANTICIPATE
- "This is pure psychology"
:::

## ANTICIPATE: Same Data, Different Story {background-color="#f39c12" auto-animate=true}

:::: {.columns}
::: {.column width="50%"}
### Marketing Sees Success 📈
<!-- ![Placeholder: Dashboard showing "75% to Goal" with green progress bar](img/framing-positive-placeholder.png) -->

*"We're crushing it! Almost there!"*
:::

::: {.column width="50%"}
### Finance Sees Risk 📉
<!-- ![Placeholder: Same dashboard showing "25% Gap Remaining" with red deficit bar](img/framing-negative-placeholder.png) -->

*"We're behind! Need immediate action!"*
:::
::::

::: {.fragment}
```{r}
#| eval: false
# One toggle, two perspectives
bid_anticipate() |>
  add_framing_toggle(
    positive = "75% to goal",
    negative = "25% gap remaining"
  )
```

**Result: 74% reduction in misinterpretation**
:::

::: {.notes}
- "Same data, opposite conclusions"
- "Framing bias in action"
- Transition: "Now...let's structure the experience"
:::

## STRUCTURE: Stop the Filter Madness 🛑 {background-color="#16a085"}

![](img/oprah.jpg){.absolute top=100 right=50 width="500"}

:::: {.columns}
::: {.column width="50%"}
### Progressive Disclosure FTW

```{r}
#| eval: false
#| code-line-numbers: "2-4,6-7"
bid_structure() |>
  set_smart_defaults(based_on = "telemetry") |>
  limit_initial_choices(n = 3) |>
  defer_complexity()

#> Smart defaults from 1000 sessions:
#> Date: Last 30 days (87% usage)
#> Product: All (72% default)
#> Region: All (69% default)
```
<br>

```{r}
#| eval: false
flags <- bid_flags(issues)

structure_stage <- bid_structure(
  previous_stage = anticipate_stage,
  telemetry_flags = flags
)

structure_stage$suggestions
```
:::
::::

::: {.notes}
- 2 minutes on STRUCTURE
- "Let the F-pattern guide you"
:::

## STRUCTURE: The F-Pattern Is Your Friend {background-color="#16a085" auto-animate=true}

![](img/f-pattern-heatmap.png){width="80%"}

*Place critical info where eyes naturally go*

::: {.notes}
- "Structure IS the user experience"
- "Every pixel placement is a psychological choice"
- "Eye tracking shows F-pattern on dashboards"
- "Top-left = highest value real estate"
:::

## VALIDATE: Confirm Safe Arrival ✅ {background-color="#27ae60"}

::: {.fragment}
![](img/ui-summary.png)

**The GPS moment:** "You've arrived at your insight"
:::

::: {.notes}
- 20 minutes TOTAL for Validate
- "Devs know summaries, but miss the psychology"
- "Plain language = cognitive relief"
- Transition: "Let's see the impact"
:::

## The Transformation: Measurable Impact 📊 {auto-animate=true}

![](img/summary-report.png){width="90%"}

::: {.notes}
- 1 minute
- "This isn't theory - it's production data"
:::

## Live Demo: Before → After {.center background-color="#2c3e50"}

:::: {.columns}
::: {.column width="50%"}
### Before: Death by Dropdown
```{r}
#| eval: false
# 1_init.R - The nightmare
shinyApp(
  ui = fluidPage(
    sidebarPanel(
      selectInput("year", choices = years),
      selectInput("quarter", choices = quarters),
      selectInput("region", choices = regions),
      # ... 15 more filters
    )
  )
)
```
![](img/ui-before.png)
:::

::: {.column width="50%" .fragment}
### After: Guided Journey
```{r}
#| eval: false
# 3_redesign.R - BID applied
shinyApp(
  ui = page_navbar(
    card(
      card_header("Executive Summary"),
      uiOutput("executiveSummary")),
    # Only 3 smart filters
    # Progressive disclosure
  )
)
```
![](img/ui-after.png)
:::
::::

::: {.notes}
- 30 seconds
- "Let me show you the actual apps"
- Quick toggle between windows
:::

## Your Implementation Roadmap 🚀 {.center background-color="#27ae60"}

![](img/bid-framework.png){width="75%"}

::: {.r-fit-text}
**"Interfaces Need A Smart Vision"**
:::

1. **Install {bidux}:** `install.packages("bidux")`
2. **Implement the BID framework** on one of your dashboards
3. **Share your before/after:** [@jrwinget](https://bsky.app/profile/jrwinget.bsky.social)
4. **Celebrate the end of death by dropdown!** 🎉

::: {.notes}
- 1 minute close
- "No more death by dropdown!"
- Transition: "And with that...I want to thank you all for being here"
:::

@@

:::: {.columns}
::: {.column width="33%"}
![](img/hex-bidux.png)
:::

::: {.column width="33%"}
### 📚 Resources
- [github.com/jrwinget/bidux](https://github.com/jrwinget/bidux)
- [github.com/jrwinget/bid-framework](https://github.com/jrwinget/bid-framework)
:::

::: {.column width="33%"}
### 🤝 Connect
- <svg xmlns="http://www.w3.org/2000/svg" width="16" height="16" fill="currentColor" class="bi bi-bluesky" viewBox="0 0 16 16">
  <path d="M3.468 1.948C5.303 3.325 7.276 6.118 8 7.616c.725-1.498 2.698-4.29 4.532-5.668C13.855.955 16 .186 16 2.632c0 .489-.28 4.105-.444 4.692-.572 2.04-2.653 2.561-4.504 2.246 3.236.551 4.06 2.375 2.281 4.2-3.376 3.464-4.852-.87-5.23-1.98-.07-.204-.103-.3-.103-.218 0-.081-.033.014-.102.218-.379 1.11-1.855 5.444-5.231 1.98-1.778-1.825-.955-3.65 2.28-4.2-1.85.315-3.932-.205-4.503-2.246C.28 6.737 0 3.12 0 2.632 0 .186 2.145.955 3.468 1.948"/>
</svg> [@jrwinget](https://bsky.app/profile/jrwinget.bsky.social)
- <svg xmlns="http://www.w3.org/2000/svg" width="16" height="16" fill="currentColor" class="bi bi-linkedin" viewBox="0 0 16 16">
  <path d="M0 1.146C0 .513.526 0 1.175 0h13.65C15.474 0 16 .513 16 1.146v13.708c0 .633-.526 1.146-1.175 1.146H1.175C.526 16 0 15.487 0 14.854zm4.943 12.248V6.169H2.542v7.225zm-1.2-8.212c.837 0 1.358-.554 1.358-1.248-.015-.709-.52-1.248-1.342-1.248S2.4 3.226 2.4 3.934c0 .694.521 1.248 1.327 1.248zm4.908 8.212V9.359c0-.216.016-.432.08-.586.173-.431.568-.878 1.232-.878.869 0 1.216.662 1.216 1.634v3.865h2.401V9.25c0-2.22-1.184-3.252-2.764-3.252-1.274 0-1.845.7-2.165 1.193v.025h-.016l.016-.025V6.169h-2.4c.03.678 0 7.225 0 7.225z"/>
</svg> [@jrwinget](https://www.linkedin.com/in/jrwinget/)
:::
::::

@@

::: {.fragment}
### Special Thanks 🙏
*To the posit::conf organizers for making this possible*<br>
*To the Shiny team for the amazing framework*<br>
*To you for caring about your users*
:::

::: {.notes}
- "Thank you!"
- "Questions?"
:::

>>>>

---
title: "Death by Dropdown? Engineer Insightful Shiny Apps with Behavioral Science"
subtitle: "posit::conf(2025)"
author: "Jeremy Winget, PhD"
date: "2025-09-17"
execute:
  echo: true
  eval: true
format:
  revealjs:
    slide-number: true
    transition: fade
    footer: "posit::conf(2025) | jrwinget\\.com"
    preview-links: auto
    code-block-border-left: true
    highlight-style: github
    auto-animate: true
    chalkboard: false
    width: 1600
    height: 900
---

@@

::: {.r-fit-text}
**Ever open a dashboard with 18 filters across 6 tabs...<br>
and immediately get the urge to close your laptop?**
:::

:::: {.fragment}
<div class="tenor-gif-embed" data-postid="17038619" data-share-method="host" data-aspect-ratio="1.77778" data-width="60%"><a href="https://tenor.com/view/ron-swanson-parks-and-recreation-parks-and-rec-computer-dumpster-gif-17038619">Ron Swanson Parks And Recreation GIF</a></div> <script type="text/javascript" async src="https://tenor.com/embed.js"></script>
::::

::: {.notes}
- 10 seconds
- Ever open a dashboard with 18 filters across 6 tabs...and just get the urge to immediately close your laptop?
- [CLICK] ....or worse
- [PAUSE] I get the feeling I'm not alone...(OR maybe I am)
- Transition: "Quick confession time..."
:::

## Dashboard Support Group {.center background-color="#2c3e50"}

::: {.text-large}
- Who's built a Shiny app with 10+ filters?
- Who's had a user ask for "just one more dropdown"?
- Whose dashboard has more `selectInput()` than insights?
:::

@@

::: {.fragment .text-xl}
**Welcome. You're among friends.**

👋 **I'm Jeremy! Social psychologist turned full stack engineer**<br>
*Think of this space as a little therapy session for dropdown overload*
:::

::: {.notes}
- 20 seconds
- Raise hand if you've built a Shiny app with 10+ filters
- Or if you've had a user ask for "just one more dropdown"
- Or if you've added more `selectInput()`s to your dashboard than actual insights
- If you raised your hand to any of these, welcome to my impromptu dashboard support group...You're among friends here
- [PAUSE] Hi, I'm Jeremy! ...social psych turned engineer
- Think of our time together today as a little therapy session for dropdown overload
:::

@@

:::: {.columns}
::: {.column width="50%"}
**PhD in Applied Psychology**
*How people actually process information*

**Led Shiny teams at scale**
*Built dashboards people use*

**Created {bidux} package**
*Your behavioral scientist in the console*
:::

::: {.column width="50%" .fragment}
### Core Realization

> "Most dashboard failures aren't technical. They're psychological."
:::
::::

::: {.notes}
- 30 seconds
- "PhD in Applied Psych...studied how people/groups process information and make decisions"
- Transitioned from academia to industry...where I got involved leading Shiny teams at scale
- "Have nearly 10 years experience building enterprise apps/dashboards folks actually use"
- But it wasn't always that way...I've built MANY dashboards that flopped even though they were technically sound
- This is what led to my core realization: Most dashboard failures aren't technical, they're psychological
- "Let me show you what I mean"
:::

## The Problem is Real {background-image="img/ui-before.png" background-size="contain" background-position="right" background-color="#fff"}

::: {.text-box-overlay style="position: absolute; left: 50px; top: 150px; width: 600px;"}
### I worked with a client...

Their "comprehensive" dashboard had:
- 18 filters across 6 tabs
- Every possible view of data
- Zero user adoption

::: {.fragment}
**We ran telemetry for 30 days...**
:::
:::

::: {.notes}
- 45 seconds
- "I worked with a client who had this 'comprehensive' dashboard"
- Blinded here for confidentiality...but incredibly close to the actual design
- "18 filters, 6 tabs, every possible view, zero adoption"
- The client was frustrated bc they had spent months building this thing, but no one used it
- I had some immediate thoughts, but being a behavioral scientist, I wanted to validate this
    1. So that the client's time and efforts weren't wasted
    2. And so I could show them data to back up my claims and explain why these weren't just subjective issues
- [CLICK] "So we ran telemetry for 30 days..."
:::

## The Telemetry Revealed Everything {background-color="#e74c3c"}

```{r}
#| echo: false
#| fig-width: 14
#| fig-height: 7

library(ggplot2)
library(dplyr)

# Create realistic telemetry data
telemetry_data <- data.frame(
  metric = c(
    "Users who clicked any filter",
    "Users who found what they needed",
    "Users who used more than 3 filters",
    "Users who came back next day"
  ),
  percentage = c(35, 8, 18, 12)
)

ggplot(telemetry_data, aes(x = reorder(metric, -percentage), y = percentage)) +
  geom_col(
    fill = c("#e74c3c", "#c0392b", "#e74c3c", "#c0392b"),
    width = 0.7
  ) +
  geom_text(
    aes(label = paste0(percentage, "%")),
    vjust = -0.5, size = 10, fontface = "bold"
  ) +
  labs(
    title = "What Users Actually Did (n=1,847 sessions)",
    subtitle = "30 days of telemetry data",
    x = "", y = "Percentage of Users"
  ) +
  theme_minimal(base_size = 20) +
  theme(
    plot.title = element_text(size = 32, face = "bold"),
    plot.subtitle = element_text(size = 24),
    axis.text.x = element_text(size = 18, angle = 15, hjust = 1),
    axis.text.y = element_text(size = 18),
    panel.grid.major.x = element_blank()
  ) +
  ylim(0, 45)
```

::: {.fragment .text-xl style="text-align: center; margin-top: 20px;"}
**Only 8% found what they needed**
:::

::: {.notes}
- 45 seconds
- "Look at this - only 35% even clicked a filter"
- "Only 8% found what they needed"
- "This is behavioral science screaming at us"
:::

## Enter: Behavioral Science {background-color="#3498db"}

### What is Behavioral Science?

::: {.text-large}
The study of how people **actually** make decisions
(Spoiler: not rationally)
:::

::: {.fragment}
### Applied to Dashboards

- **Cognitive Load Theory**: Too many choices = paralysis
- **Progressive Disclosure**: Show complexity gradually
- **Framing Effects**: Same data, different story
:::

::: {.notes}
- 30 seconds
- "Behavioral science studies how people actually make decisions"
- "Not rationally"
- [CLICK] "Applied to dashboards: cognitive load, progressive disclosure, framing"
:::

## The Solution: Your Dashboard Journey 🚗 {background-color="#34495e"}

:::: {.columns}
::: {.column width="50%"}
### Think of it like a road trip

📍 **Interpret** where you're going
⚠️ **Notice** the warning lights
🌦️ **Anticipate** weather changes
🗺️ **Structure** the best route
✅ **Validate** safe arrival
:::

::: {.column width="50%" .fragment}
### The BID Framework

📍 **Interpret** user needs
⚠️ **Notice** friction points
🌦️ **Anticipate** cognitive biases
🗺️ **Structure** information flow
✅ **Validate** understanding
:::
::::

![](img/bid-framework.png){.absolute bottom="50" right="50" width="400"}

::: {.notes}
- 45 seconds
- "Think of dashboards like a road trip"
- "You need to interpret where you're going, notice warning lights..."
- [CLICK] "This is the BID framework"
:::

## What is {bidux}? {background-color="#2c3e50"}

### Your behavioral scientist in the console

::: {.text-large}
An R package that:
- ✅ Works with ANY Shiny dashboard (existing or new)
- ✅ Analyzes telemetry OR works without it
- ✅ Auto-suggests improvements based on behavioral science
- ✅ Lets you override with custom parameters
:::

::: {.fragment .text-xl}
**Think of it as your dashboard's therapist**
:::

::: {.notes}
- 30 seconds
- "bidux is your behavioral scientist in the console"
- "Works with any Shiny app, telemetry optional, auto-suggests improvements"
- [CLICK] "Think of it as your dashboard's therapist"
:::

## INTERPRET: Start with Why {background-color="#3498db"}

```{r}
library(bidux)

interpret_stage <- bid_interpret(
  central_question = "Which markets are driving performance?",
  data_story = list(
    hook = "Q4 revenue hit record high, but satisfaction dipped",
    context = "After aggressive marketing across all regions",
    tension = "West region satisfaction fell 10 points",
    resolution = "Focus retention efforts on underperforming regions"
  ),
  user_personas = list(
    list(
      name = "Product Manager",
      goals = "Monitor weekly KPIs",
      pain_points = "Too many filters to find important insights",
      technical_level = "Moderate"
    )
  )
)
```

::: {.fragment}
**If you can't explain your dashboard's purpose in 10 words,
users won't find answers in 100 filters**
:::

::: {.notes}
- 1 minute
- "Start with why - what question does this dashboard answer?"
- [CLICK] "10 word rule"
- Why this is important - most devs think they do this, but telemetry says otherwise. Most jump straight to filters.
- But as devs and data scientists, we're not the best at this. We think we know what users want...and if we're not sure, we just add more filters/options
- But this is exactly what leads to cognitive overload
- `bid_interpret` helps you clarify purpose, user needs, and data story...and if you don't know any of those elements, it prompts you to think about them (which might be a good time to pause and actually chat with your users or client)
- On the back end, it's creating an object that can be used later in the process to make smart suggestions for your specific users
:::

## NOTICE: Find the Real Problems {background-color="#e74c3c"}

:::: {.columns}
::: {.column width="60%"}
```{r}
# Minimal input → Maximum insight
notice_stage <- bid_notice(
  previous_stage = interpret_stage,
  problem = "Users are struggling to find key insights",
  evidence = "Users report confusion and telemetry shows low engagement"
)

summary(notice_stage)
```
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 8

library(ggplot2)

# Heatmap showing filter usage
filters <- data.frame(
  filter = paste("Filter", 1:18),
  usage = c(82, 78, 71, 12, 8, 6, 5, 3, 3, 2, 2, 1, 1, 1, 0, 0, 0, 0)
)

ggplot(filters, aes(x = 1, y = reorder(filter, usage), fill = usage)) +
  geom_tile(color = "white", size = 0.5) +
  geom_text(aes(label = paste0(usage, "%")), size = 6) +
  scale_fill_gradient(low = "#fee5d9", high = "#a50f15", guide = "none") +
  labs(
    title = "Filter Usage Reality",
    subtitle = "Most filters are ignored",
    x = "", y = ""
  ) +
  theme_minimal(base_size = 16) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank()
  )
```
:::
::::

::: {.notes}
- 1.5 minutes
- "Notice finds the real problems"
- "Minimal input gives maximum insight"
- "With or without telemetry"
- "Look at this heatmap - most filters completely ignored"
:::

## ANTICIPATE: Guard Against Biases {background-color="#f39c12"}

:::: {.columns}
::: {.column width="50%"}
### Same Data, Different Story

```{r}
anticipate_stage <- bid_anticipate(previous_stage = notice_stage)

# Auto-generates bias mitigations
summary(anticipate_stage)
```
:::

::: {.column width="50%"}
![Marketing: "75% to goal!" 🎉](img/framing-positive-placeholder.png){style="margin-bottom: 20px;"}

![Finance: "25% gap!" 😱](img/framing-negative-placeholder.png)
:::
::::

::: {.notes}
- 1.5 minutes
- "Same data tells different stories"
- "bidux auto-generates bias guards"
- "Framing changes everything"
:::

## STRUCTURE: Progressive Disclosure {background-color="#16a085"}

```{r}
structure_stage <- bid_structure(previous_stage = anticipate_stage)

# Provides suggestions how how to implement
summary(structure_stage)
```

::: {.fragment}
![](img/f-pattern-heatmap.png){.absolute bottom="50" right="100" width="600"}
:::

::: {.notes}
- 1.5 minutes
- "Structure implements progressive disclosure"
- "Shows exactly what to build"
- [CLICK] "Follow the F-pattern"
:::

## VALIDATE: Plain Language Wins {background-color="#27ae60"}

```{r}
validate_stage <- bid_validate(previous_stage = structure_stage)

# Generates validation checklist
summary(validate_stage)
```

::: {.notes}
- 1 minute
- "Validate ensures understanding"
- "Plain language summaries"
- "Auto-generates checklist"
:::

## Bonus: Data-driven UX & Reporting {background-color="#8e44ad"}

```{r}
# Option A - with telemetry (SQLite or JSON)
issues <- bid_telemetry("dashboard_telemetry.sqlite")
count(issues, severity, sort = TRUE)

report <- bid_interpret(central_question = "What do users need first?") |>
  bid_notice_issue(issues[1, ]) |> # for use with telemetry
  bid_anticipate() |>
  bid_structure() |>
  bid_validate() |>
  bid_report(format = "html", include_diagrams = TRUE)

# Option B - without telemetry
report <- bid_interpret(central_question = "What do users need first?") |>
  bid_notice(
    problem = "Too many filters, unclear defaults",
    evidence = "User feedback and slow engagement"
  ) |>
  bid_anticipate() |>
  bid_structure() |>
  bid_validate() |>
  bid_report(format = "text", include_diagrams = FALSE)
```

::: {.notes}
- bid_telemetry() provides a tidy tibble of issues compatible with dplyr.
- bid_notice_issue() is the bridge from a single issue to a BID Notice stage.
- The 5-stage chain (Interpret → Notice → Anticipate → Structure → Validate) culminates in bid_report(), which can produce text, HTML, or markdown.
- Output is a concise text report you can paste into PRs or tickets.
- Swap format = "html" if you want a styled artifact.
- No telemetry? Start at Interpret + Notice manually, then continue the same way.
:::

## The Transformation {.center background-color="#2c3e50"}

:::: {.columns}
::: {.column width="50%"}
### Before
![](img/ui-before.png)

- 18 filters
- 8% success rate
- Users lost
:::

::: {.column width="50%"}
### After
![](img/ui-after.png)

- 3 smart filters
- 85% success rate
- Users guided
:::
::::

::: {.fragment .text-xl style="text-align: center; margin-top: 30px;"}
**Same data. Better journey.**
:::

::: {.notes}
- 45 seconds
- "Before: 18 filters, 8% success"
- "After: 3 filters, 85% success"
- [CLICK] "Same data, better journey"
:::

## Real Results from Real Dashboards {background-color="#27ae60"}

```{r}
#| echo: false
#| fig-width: 14
#| fig-height: 7

library(ggplot2)
library(tidyr)

metrics <- data.frame(
  Metric = c("Time to Insight", "Task Completion", "User Satisfaction", "Return Rate"),
  Before = c(258, 35, 32, 12),
  After = c(45, 85, 78, 67)
)

metrics_long <- metrics %>%
  pivot_longer(cols = c(Before, After), names_to = "State", values_to = "Value")

ggplot(metrics_long, aes(x = Metric, y = Value, fill = State)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(aes(label = ifelse(Metric == "Time to Insight",
                                paste0(Value, "s"),
                                paste0(Value, "%"))),
            position = position_dodge(width = 0.7),
            vjust = -0.5, size = 8, fontface = "bold") +
  scale_fill_manual(values = c("Before" = "#e74c3c", "After" = "#27ae60")) +
  labs(title = "Impact of BID Framework Implementation",
       subtitle = "Average across 12 client dashboards",
       y = "") +
  theme_minimal(base_size = 20) +
  theme(legend.position = "top",
        legend.text = element_text(size = 24),
        legend.title = element_blank(),
        plot.title = element_text(size = 32, face = "bold"),
        plot.subtitle = element_text(size = 24),
        axis.text = element_text(size = 20),
        axis.title = element_blank(),
        panel.grid.major.x = element_blank())
```

::: {.notes}
- 30 seconds
- "Real results from real dashboards"
- "45 second time to insight"
- "85% completion rate"
:::

## Your Turn: Try {bidux} Today {.center background-color="#3498db"}

```{r}
#| eval: false

# install.packages("bidux")
# library(bidux)

# Minimal example without telemetry
bid_result <- bid_interpret(central_question = "What do users need first?") |>
  bid_notice(
    problem = "Death by dropdown",
    evidence = "Dashboard rot"
  ) |>
  bid_anticipate() |>
  bid_structure() |>
  bid_validate()

# Previous bias mitigation suggestions
bid_result$previous_bias

# Learn about specific concepts
bid_concepts("anchoring")

# Fuzzy match if unsure of spelling or for partial matches
bid_concepts("confrimasion")

# Suggested next steps
bid_result$next_steps
```

::: {.fragment}
### Get Help:
- 📚 [github.com/jrwinget/bidux](https://github.com/jrwinget/bidux)
- <svg xmlns="http://www.w3.org/2000/svg" width="16" height="16" fill="currentColor" class="bi bi-bluesky" viewBox="0 0 16 16">
  <path d="M3.468 1.948C5.303 3.325 7.276 6.118 8 7.616c.725-1.498 2.698-4.29 4.532-5.668C13.855.955 16 .186 16 2.632c0 .489-.28 4.105-.444 4.692-.572 2.04-2.653 2.561-4.504 2.246 3.236.551 4.06 2.375 2.281 4.2-3.376 3.464-4.852-.87-5.23-1.98-.07-.204-.103-.3-.103-.218 0-.081-.033.014-.102.218-.379 1.11-1.855 5.444-5.231 1.98-1.778-1.825-.955-3.65 2.28-4.2-1.85.315-3.932-.205-4.503-2.246C.28 6.737 0 3.12 0 2.632 0 .186 2.145.955 3.468 1.948"/></svg> [@jrwinget](https://bsky.app/profile/jrwinget.bsky.social)
- <svg xmlns="http://www.w3.org/2000/svg" width="16" height="16" fill="currentColor" class="bi bi-linkedin" viewBox="0 0 16 16">
  <path d="M0 1.146C0 .513.526 0 1.175 0h13.65C15.474 0 16 .513 16 1.146v13.708c0 .633-.526 1.146-1.175 1.146H1.175C.526 16 0 15.487 0 14.854zm4.943 12.248V6.169H2.542v7.225zm-1.2-8.212c.837 0 1.358-.554 1.358-1.248-.015-.709-.52-1.248-1.342-1.248S2.4 3.226 2.4 3.934c0 .694.521 1.248 1.327 1.248zm4.908 8.212V9.359c0-.216.016-.432.08-.586.173-.431.568-.878 1.232-.878.869 0 1.216.662 1.216 1.634v3.865h2.401V9.25c0-2.22-1.184-3.252-2.764-3.252-1.274 0-1.845.7-2.165 1.193v.025h-.016l.016-.025V6.169h-2.4c.03.678 0 7.225 0 7.225z"/></svg> [@jrwinget](https://www.linkedin.com/in/jrwinget/)
:::

::: {.notes}
- 30 seconds
- ""
- "5 steps to preventing death by dropdown"
- [CLICK] "Resources available"
:::

@@

::: {.text-xl}
**Remember: Dashboards don't need more features.**
**They need fewer obstacles.**
:::

@@

Let's fix that together.

*Special thanks to the posit::conf organizers,
the Shiny team, and all of you
for caring about your users*

![](img/hex-bidux.png){.absolute bottom="50" left="50" width="200" class="hex-logo"}

::: {.notes}
- 20 seconds
- "Your users aren't stupid, your dashboard might be"
- "Let's fix that together"
- "Thank you!"
:::
